Stage-Splitting a Modal Language

### 2014-09-30 paper/semantics.tex
Fixes from meeting with Kayvon.

\lang\ is a typed two-stage lambda calculus with products, sums, and isorecursive types.
Its statics are adapted from the linear temporal logic given by \cite{davies96},
restricted to two stages and extended with general sums and recursion.%
\footnote{This restriction to two stages is made primarily for simplicity of
presentation. All of these techniques could feasibly be extended to more
complicated stage systems.}
%The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of 
%, as language to define and analyze the splitting transform.

%The following subsections provide a brief overview of the type system for \lang,
%and highlight important differences in its semantics from this prior work.

@@

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
a stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
monostaged language called \langTwo. Then, at stage~\bbtwo, we complete
evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
rules for this judgment are not shown, but they are standard.)

@@

If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the stage \bbone\ result would be
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s and $\next$s cancel to yield the residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20|
twice! To avoid this duplication, instead of treating \verb|next {f 20}| as a
value and substituting it into the body of the \verb|let|, we instead bind
\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
variable.

@@

Our dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and $\reifysym$ (reification of substitutions into let
bindings).

The algorithm described above creates two different kinds of expressions which
cannot be evaluated further at stage \bbone:
\begin{itemize}
\item Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."

@@

>>>>

\lang\ is a typed two-stage lambda calculus with products, sums, and isorecursive types.
Its statics are adapted from the linear temporal logic given by \cite{davies96},
restricted to two stages and extended with general sums and recursion.%
\footnote{This restriction to two stages is made primarily for simplicity of
presentation. All of these techniques could feasibly be extended to more
complicated stage systems.} 
Although \cite{davies96} was mostly interested in describing partial evaluation,
the system is equally apt as an input to splitting.

@@

Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
a stage \bbtwo\ term with no stage~\bbone\ content, 
which can be described as a term in a monostaged language called \langTwo. 
Then, at stage~\bbtwo, we complete
evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
rules for this judgment are not shown, but they are standard.)

@@

If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the stage \bbone\ result would be
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s and $\next$s cancel to yield the residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.
Accordingly, instead of treating \verb|next {f 20}| as a
value and substituting it into the body of the \verb|let|, we instead bind
\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
variable.

@@

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and $\reifysym$ (reification of substitutions into let
bindings).

\lang\ creates two different kinds of expressions which
cannot be evaluated further at stage \bbone:
\begin{itemize}
\item Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.

@@

\TODO [below here will get moved to related work]
